---
output:
  html_document: default
  pdf_document: default
---

## Session 3: EURODOL (non-Seasonal) ARIMA(p,d,q) Modelling

### Aim: To identify (and fit) at least 2 plausible non-Seasonal ARIMA(p,d,q) models for the EURODOL time series

#### EURODOL: Exchange rate; US dollars per euro or ECU (USD / EUR). January 1995 - December 2019; 25 years. Data source: official web Banco de España (https://www.bde.es/webbde/es/estadis/infoest/tipos/tipos.html).

### I. Exploratory data analysis

Import the data

```{r}
serie=ts((read.table("eurodol.dat")),start=1995,freq=12)
print(round(serie,3))
```

Plot the time series EURODOL; describe it…

```{r}
plot(serie)
abline(v=1995:2020,col=4,lty=3)


```

Remarks…

Mention/describe here all known events in the period of study that might have influenced the EURODOL time series: …

### II. Transformation into stationarity

**Is Variance constant?**

```{r}
boxplot(serie~floor(time(serie)))
#Mean-variance plot
m=apply(matrix(serie,ncol=25),2,mean)
v=apply(matrix(serie,ncol=25),2,var)
plot(v~m,main="Mean-Variance plot")
```

We consider the varianza to be constant! A log transformation not needed: it does not change variability pattern.

**Is seasonality present?**


```{r}
monthplot(serie)
```



Nope! It does not make sense to have seasonality in an exchange-rate time series variable!

**Is the mean constant?**

Plot again the whole time series

```{r}
plot(serie, ylab = "USD/EUR")
abline(h=mean(serie),col=2)
```

Hmmm, not clear! Let us take a look at the ACF:

```{r}
acf(serie,ylim=c(-1,1),lag.max=60,col=c(2,rep(1,11)))
```


The ACF decays very slowly towards zero, suggesting non-stationarity (non-constant mean yet). At this point we need to take a decision:

- **Case 1.** Maybe not constant mean yet: we need to make it stationary by taking regular differences, or

- **Case 2.** Already constant with non-zero mean = 1.21 (reversion to the mean): assume that the EURODOL series is already stationary!!

#### **Case 1.** The EURODOL series is assumed to be non-stationary: Make the series stationary. Take a regular differentiation: d=1.

```{r}
d1serie=diff(serie)
plot(d1serie)
abline(h=0)
```


Now, the mean seemingly constant at zero!!!

**Is an extra differentiation needed? Take an extra differentiation and compare the variances.**

```{r}
var(serie,na.rm=T)
var(d1serie,na.rm=T)
var(diff(d1serie),na.rm=T)
```

Nope! We observe that with extra regular difference the variance increases…

**Is the current series already stationary? Plot the P(ACF) and decide!!**

```{r}
par(mar = c(3,4,3,1))
acf(d1serie,ylim=c(-1,1),lag.max=60,col=c(2,rep(1,11)),na.action=na.pass, main="Series d1serie")
```


Conclusion: The differenced (d=1) exchange rate time series EURODOL (d1serie) is stationary: constant mean(=0); constant variance. The covariance structure only depends on the lags and not on the time origen.

In this case, d1serie=$W_t=(1−B)X_t$ with $μ_{W_t}=0$.

#### **Case 1.** The EURODOL series is assumed to be already stationary

In this case, serie=$W_t=X_t$ with $μ_{W_t}=1.21$.


### III. Model Identification

#### Inspect the P(ACF) and propose >=2 models (principle of parsimony!)

#### **Case 1**. Identify possible values of p and q for the series **d1serie**

```{r}
par(mfrow=c(1,2), mar = c(3,4,3,1))
acf(d1serie,ylim=c(-1,1),lag.max=60,col=c(2,rep(1,11)),na.action=na.pass, main="Series d1serie")
pacf(d1serie,ylim=c(-1,1),lag.max=60,col=c(rep(1,11),2),na.action=na.pass)
```


#### Two models proposed for the stationary time series **d1serie** (W_t). Model 1: AR(1) and Model 2: MA(1).

#### **Case 2**. Identify possible values of p and q for the series: serie

Reminder: plot the time series

```{r}
plot(serie)
abline(v=1995:2020,col=4,lty=3)
abline(h=1)
abline(h=mean(serie), col=2)
```

ACF and PACF of **serie**

```{r}
par(mfrow=c(1,2))
acf(serie,ylim=c(-1,1),lag.max=60,col=c(2,rep(1,11)))
pacf(serie,ylim=c(-1,1),lag.max=60,col=c(rep(1,11),2))
par(mfrow=c(1,1))
```

We observe an ACF that decays slowly towards zero! BUT, we force an stationary model: thus, and AR(2) identified (Model 3).

### IV. Model Estimation and validation

#### ‘Load’ the function ‘validation’: used later

```{r}
#################Validation#################################
validation=function(model,dades){
  s=frequency(get(model$series))
  resid=model$residuals
  par(mfrow=c(2,2),mar=c(3,3,3,3))
  #Residuals plot
  plot(resid,main="Residuals")
  abline(h=0)
  abline(h=c(-3*sd(resid),3*sd(resid)),lty=3,col=4)
  #Square Root of absolute values of residuals (Homocedasticity)
  scatter.smooth(sqrt(abs(resid)),main="Square Root of Absolute residuals",
                 lpars=list(col=2))
  
  #Normal plot of residuals
  qqnorm(resid)
  qqline(resid,col=2,lwd=2)
  
  ##Histogram of residuals with normal curve
  hist(resid,breaks=20,freq=F)
  curve(dnorm(x,mean=mean(resid),sd=sd(resid)),col=2,add=T)
  
  
  #ACF & PACF of residuals
  par(mfrow=c(1,2))
  acf(resid,ylim=c(-1,1),lag.max=60,col=c(2,rep(1,s-1)),lwd=1)
  pacf(resid,ylim=c(-1,1),lag.max=60,col=c(rep(1,s-1),2),lwd=1)
  par(mfrow=c(1,1))
  
  #ACF & PACF of square residuals 
  par(mfrow=c(1,2))
  acf(resid^2,ylim=c(-1,1),lag.max=60,col=c(2,rep(1,s-1)),lwd=1)
  pacf(resid^2,ylim=c(-1,1),lag.max=60,col=c(rep(1,s-1),2),lwd=1)
  par(mfrow=c(1,1))
  
  #Ljung-Box p-values
  par(mar=c(2,2,1,1))
  tsdiag(model,gof.lag=7*s)
  cat("\n--------------------------------------------------------------------\n")
  print(model)
  
  #Stationary and Invertible
  cat("\nModul of AR Characteristic polynomial Roots: ", 
      Mod(polyroot(c(1,-model$model$phi))),"\n")
  cat("\nModul of MA Characteristic polynomial Roots: ",
      Mod(polyroot(c(1,model$model$theta))),"\n")
  
  #Model expressed as an MA infinity (psi-weights)
  psis=ARMAtoMA(ar=model$model$phi,ma=model$model$theta,lag.max=36)
  names(psis)=paste("psi",1:36)
  cat("\nPsi-weights (MA(inf))\n")
  cat("\n--------------------\n")
  print(psis[1:20])
  plot(psis[1:20], pch=16, main ="Psis-weights")
  axis(1, 1:20)
  
  #Model expressed as an AR infinity (pi-weights)
  pis=-ARMAtoMA(ar=-model$model$theta,ma=-model$model$phi,lag.max=36)
  names(pis)=paste("pi",1:36)
  cat("\nPi-weights (AR(inf))\n")
  cat("\n--------------------\n")
  print(pis[1:20])
  plot(pis[1:20], pch=16, main ="Pi-weights")
  axis(1, 1:20)
  
  ##Shapiro-Wilks Normality test
  print(shapiro.test(resid(mod)))
  
  #Sample ACF vs. Teoric ACF
  par(mfrow=c(2,2),mar=c(3,3,3,3))
  acf(dades, ylim=c(-1,1) ,lag.max=36,main="Sample ACF")
  
  plot(ARMAacf(model$model$phi,model$model$theta,lag.max=36),ylim=c(-1,1), 
       type="h",xlab="Lag",  ylab="", main="ACF Teoric")
  abline(h=0)
  
  #Sample PACF vs. Teoric PACF
  pacf(dades, ylim=c(-1,1) ,lag.max=36,main="Sample PACF")
  
  plot(ARMAacf(model$model$phi,model$model$theta,lag.max=36, pacf=T),ylim=c(-1,1),
       type="h", xlab="Lag", ylab="", main="PACF Teoric")
  abline(h=0)
  par(mfrow=c(1,1))
}
################# Fi Validacio #################################
```


### **Case 1.** Estimation and Validation

#### Model 1


```{r}
(mod=arima(d1serie,order=c(1,0,0))) #Stationary AR(1)
(mod=arima(serie,order=c(1,1,0)))   #Non-stationary ARIMA(1,1,0) model for serie
```

```{r}
dades=d1serie
model=mod
validation(model,dades)
```

**Model 1:** ARIMA(1,1,0) $(1-0.2993B)(1-B)X_t=Z_t \qquad Z_t\sim WN(0,\sigma_Z^2=0.0006723)$


#### Model 2

```{r}
(mod=arima(d1serie,order=c(0,0,1))) #Stationary MA(1) model for d1serie
(mod=arima(serie,order=c(0,1,1))) #Non-stationary ARIMA(0,1,1) model for serie
```

```{r}
dades=d1serie
model=mod
validation(model,dades)
```

**Model 2:** ARIMA(0,1,1) $(1-B)X_t=(1+0.3252B)Z_t \qquad Z_t\sim WN(0,\sigma_Z^2=0.000666)$



### **Case 2.** Estimation and Validation

#### Model 3


```{r}
#Stationary AR(2) model for serie; intercept <> 0
(mod=arima(serie,order=c(2,0,0)))
```


```{r}
dades=serie
model=mod
validation(model,dades)
```

**Model 3:** ARIMA(2,0,0) $(1-1.2868B+0.3084B^2)(X_t-1.1965)=Z_t \qquad Z_t\sim WN(0,\sigma_Z^2=0.000661)$


### V. Sum Up Adequacy Measures

To do: Write down a summary of the adequacy measures

#### Model chosen? Justify your answer!

Recall:

Model 1: $(1-0.2993B)(1-B)X_t=Z_t \Rightarrow (1-1.2993B+0.2993B^2)X_t=Z_t$

Model 3: $(1-1.2868B+0.3084B^2)(X_t-1.1965)=Z_t \Rightarrow (1-1.2868B+0.3084B^2)X_t=0.0258444+Z_t$