---
title: 'Session 10: States and Parameters Estimation via Kalman Filter + Newton Raphson/EM'
Date: "4/30/2020"
output:
  html_document: default
  pdf_document: default
  word_document: default
---
#  Global warming

Data come from the website:

http://data.giss.nasa.gov/gistemp/

The GISS Surface Temperature Analysis (GISTEMP) is an estimate of global surface temperature change. Graphs and tables are updated around the middle of every month using current data files from NOAA GHCN v3 (meteorological stations), ERSST v5 (ocean areas), and SCAR (Antarctic stations), combined as described in our December 2010 publication (Hansen et al. 2010).

### Read data

```{r}
gtemp1=ts(c(t(as.matrix(read.csv2("GLB.Ts.csv")[,-1]))),start=c(1880,1),freq=12)
gtemp2=ts(c(t(as.matrix(read.csv2("GLB.Ts+dSST.csv")[,-1]))),start=c(1880,1),freq=12)

gtemp1=window(gtemp1,start=1910,end=2018)
gtemp2=window(gtemp2,start=1910,end=2018)

```

```{r}
plot(gtemp1,ylim=c(-1,1.5))
lines(gtemp2,col=2)
legend("topleft",col=c(1:2),lty=1,lwd=2,legend=c("GLB.Ts","GLB.Ts+dSST"))

```


```{r}
plot(gtemp1,gtemp2,xlab="GLB.Ts",ylab="GLB.Ts+dSST",xlim=c(-1,2),ylim=c(-1,2))
abline(c(0,1))
```

```{r}
library(astsa)
```

## A) Non-linear Optimization

```{r}
y=matrix(cbind(gtemp1,gtemp2),nc=2)
A=array(1,dim=c(2,1,nrow(y)))
mu0=0
Sigma0=10
Phi=1
cQ=0.1
cR=0.1*diag(2)

Linn=function(par){
  Phi=par[1]
  cQ=par[2]
  cR=t(matrix(c(par[3],0,0,par[4]),ncol=2))
  kf=Kfilter(y, A, mu0, Sigma0, Phi, cQ, cR)
  kf$like
}
```
```{r}
initpar=(c(1,0.1,0.1,0.1))
(sol=optim(initpar,Linn,NULL,method="BFGS",hessian=TRUE,
           control=list(maxit = 800, reltol = 1e-4)))
stderr=sqrt(diag(solve(sol$hessian)))
cbind(sol$par,stderr)
```

```{r}
(Phi=sol$par[1])
cQ=sol$par[2]
cR=matrix(c(sol$par[3],0,0,sol$par[4]),ncol=2)
(t(cQ)%*%cQ)
(S=t(cR)%*%cR)

kf=Ksmooth(y, A, mu0, Sigma0, Phi, cQ, cR)

pre=ts(c(unlist(kf$Xp)),start=c(1910,1),freq=12)
fil=ts(c(unlist(kf$Xf)),start=c(1910,1),freq=12)
smo=ts(c(unlist(kf$Xs)),start=c(1910,1),freq=12)

xlim1=c(1910,2018)
plot(gtemp1,ylim=c(-1.5,2),xlim=xlim1)
lines(gtemp2,col=2)
lines(smo,col=4,lwd=2)
abline(h=0)
```

## B) EM algorithm

```{r}
num=length(gtemp1)
y=matrix(cbind(gtemp1,gtemp2),nc=2)
A=array(1,dim=c(2,1,nrow(y)))
mu0=0
Sigma0=100
Phi=1
cQ=0.05
cR=diag(c(0.1,0.1))
(sol2=EM(y, A, mu0, Sigma0, Phi, cQ, cR, max.iter = 800, tol = 1e-8))


Phi=sol2$Phi
cQ=chol(sol2$Q)
cR=chol(sol2$R)
mu0=sol2$mu0
Sigma0=sol2$Sigma0
kf=Ksmooth(y, A, mu0, Sigma0, Phi, cQ, cR)

pre=ts(c(unlist(kf$Xp)),start=c(1910,1),freq=12)
fil=ts(c(unlist(kf$Xf)),start=c(1910,1),freq=12)
smo=ts(c(unlist(kf$Xs)),start=c(1910,1),freq=12)

par(mfrow=c(2,1),mar=c(3,3,1,1))
xlim1=c(1910,2018)
plot(gtemp1,ylim=c(-1.5,2),xlim=xlim1)
lines(smo,col=4,lwd=2)
plot(gtemp2,ylim=c(-1.5,2),xlim=xlim1,col=2)
lines(smo,col=4,lwd=2)
```


