---
title: 'Session 12. Structural Time Series \break Local Trend + non Constant Seasonality'
output:
  html_document: default
  pdf_document: default
  word_document: default
Author: LA + JAS   
---

### Data Description
#### Salary in GDP in Spain (millions of euros); 1995-2016
Data come from the website: http://www.ine.es/  Economia / Cuentas Economicas / Contabilidad Nacional Trimestral  / Contabilidad Nacional Trimestral. Base 2010. Series desde el trimestre 1/1995 hasta el ultimo trimestre publicado / PIB pm Rentas (Precios corrientes) / remuneraci?n de asalariados

### Read the two quarterly time series (the original one and the seasonally adjusted)
```{r}
serieTot=ts(read.table("salarios.dat"),start=1995,freq=4) #not updated, covers time window 1995-2016
plot(serieTot)
```

### Work with the original series, transformed into logarithm
```{r}
serie=log(serieTot[,1])
plot(serie)
```

### Seasonality analysis
```{r}
monthplot(serie)
```

### Decomposition of the time series
```{r}
plot(decompose(serie))
```

### Load the package astsa
```{r}
library(astsa)
```

## Estimation of the local trend + non-constant seasonality model
```{r}
ser=serie
N=length(ser)
y=matrix(ser,ncol=1)
A=matrix(c(1,0,1,0,0),nr=1)
Phi=matrix(c(1,0,0,0,0,1,1,0,0,0,0,0,-1,1,0,0,0,-1,0,1,0,0,-1,0,0),ncol=5)
Phi

Linn=function(para){
  cQ1=para[1]    
  cQ2=para[2]   
  cQ3=para[3]   
  cR=para[4]  
  cQ=diag(c(cQ1,cQ2,cQ3,0,0))
  kf=Kfilter(y,A,mu0,Sigma0,Phi,cQ,cR)
  return(kf$like)
}

mu0=c(y[1],rep(0,4))
Sigma0=diag(rep(0.1,5))

initpar=c(0.01,0.01,0.01,0)

est=optim(initpar,Linn,NULL,method="BFGS",hessian=TRUE)
stderr=sqrt(diag(solve(est$hessian)))
est      
cbind(est$par,stderr)
```

```{r}
para=est$par
cQ1=para[1]   # sigma_w1
cQ2=para[2]   # sigma_w2
cQ3=para[3]   # sigma_w3
cR=para[4]    # sigma de R
cQ=diag(c(cQ1,cQ2,cQ3,0,0))
```

```{r}
## variance-covariance matrix for the transition error
(Q=t(cQ)%*%cQ)
## variance for the observation error
(R=t(cR)%*%cR)

c(diag(Q)[1:3],R)

#Que fa?
(m<-StructTS(ser))

fitted(m)
plot(fitted(m))
```

### Kalman Filter to obtain the smoothing estimation of the components
```{r}
kf=Ksmooth(y,A,mu0,Sigma0,Phi,cQ,cR)   

smo=ts(data.frame(level=unlist(kf$Xs)[1+(0:87)*5],slope=unlist(kf$Xs)[2+(0:87)*5],seasonal=unlist(kf$Xs)[3+(0:87)*5]),start=1995,freq=4)
plot(smo)
```
### Extraction of the trend
```{r}
par(mar=c(3,3,1,1),mfrow=c(2,1))
plot(smo[,"level"]+smo[,"seasonal"],col=1)
lines(smo[,"level"],col=2)
plot(smo[,"slope"])
abline(h=0)
```
```{r}
plot(stl(serie,s.window=4))

```

### Comparison with the standard procedure
```{r}
modstl=stl(serie,s.window=4)
par(mfrow=c(2,1),mar=c(3,3,1,1))

plot(modstl$time.series[,2]+modstl$time.series[,3])
lines(smo[,"level"],col=2)

plot(stl(serie,s.window=4)$time.series[,1])
lines(smo[,"seasonal"],col=2)

par(mfrow=c(1,1))
cbind(stl=stl(serie,s.window=4)$time.series[,2]+stl(serie,s.window=4)$time.series[,3],smo[,"level"])

```
### Seasonally adjusted series comparison
```{r}
serofi=serieTot[,2]
serkf=exp(y-smo[,"seasonal"])
serstl=exp(modstl$time.series[,2]+modstl$time.series[,3])
plot(serofi)
lines(serkf,col=2)
lines(serstl,col=4)
legend("topleft",legend=c("Ofi","KF","STL"),lty=1,col=c(1,2,4))

```
```{r}
serofi=serieTot[,2]
serkf=exp(y-smo[,"seasonal"])
serstl=exp(modstl$time.series[,2]+modstl$time.series[,3])
plot(serofi,xlim=c(2010,2017),ylim=c(115,140),type="o")
lines(serkf,col=2)
lines(serstl,col=4)
legend("bottomleft",legend=c("Ofi","KF","STL"),lty=1,col=c(1,2,4))

```
```{r}
cbind(serofi,serkf,serstl)

pairs(as.data.frame(cbind(serofi,serkf,serstl)),lower.panel=NULL)


```






