--
title: "Assignment 3 Statistical Learning"
author: "Miguel Benítez, Iker Caballero, Alejandro Peraza"
date: "2023-02-26"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(rsample)
library(pROC)
library(caret)
```

# Comparing discriminant rules. ROC curve and other methods

## Data loading and splitting
```{r}
spam <- read.table("spambase.data",sep=",")
spam.names <- c(read.table("spambase.names",sep=":",skip=33,nrows=53,as.is=TRUE)[,1],
                "char_freq_#",read.table("spambase.names",sep=":",skip=87,nrows=3,as.is=TRUE)[,1],"spam.01")

names(spam) <- spam.names 
n <- dim(spam)[1]
p <- dim(spam)[2]-1

View(spam)

set.seed(1726)
split <- rsample::initial_split(spam, prop = 0.7, strata = spam.01)
training_set <- rsample::training(split)
testing_set <- rsample::testing(split)
X_train <- scale(training_set[,1:p]) # Scale the predictors
y_train <- training_set[,(p + 1)]
X_test <- scale(testing_set[,1:p]) # Scale the new data
y_test <- testing_set[,(p + 1)]

length(y_test)
length(y_train)
```

## Logistic regression fitted by maximum likelihood

In order to estimate the logistic regression model, we use the "glm" package:

```{r}
logit_mod <- glm(y_train ~ X_train,family = "binomial")
test_logit <- predict(logit_mod,X_test,type="response")
```

Now, we can get the ROC curve using the "pROC" package:

```{r}
plot(roc(y_test~test_logit),col="blue")
```

For comparison purposes, we can also show the missclasification rate and the lval measure:

```{r}
predict = ifelse(test_logit > 0.5,1,0)
accuracy = caret::confusionMatrix(factor(predict),factor(y_test),positive="1")$overall[1]
lval = (1/n)*colSums(y_test*log(test_logit)+(1-y_test)*log(1-test_logit),na.rm = TRUE)
```

## Logistic regression fitted by Lasso

(Parte de Miguel)

```{r}
# Use the training sample to fix the tunning parameters (when needed)
# and to estimate the model parameters (when needed)
```

(Diría de usar todos la librería pROC)

```{r}
# Use the test sample to compute and plot the ROC curve for each rule.
## Store the ROC curve plot in a variable to display them at the end
# Compute also the misclassification rate for each rule when using the
# cut point c = 1/2.
## Store the ROC curve plot in another variable to display them at the end
```

```{r}
# Compute l_val for each rule.
## Store it in a variable to display at the end
```

## k-nn binary regression

(Parte de Alejandro)

```{r}
# Use the training sample to fix the tunning parameters (when needed)
# and to estimate the model parameters (when needed)
```

(Diría de usar todos la librería pROC)

```{r}
# Use the test sample to compute and plot the ROC curve for each rule.
## Store the ROC curve plot in a variable to display them at the end
# Compute also the misclassification rate for each rule when using the
# cut point c = 1/2.
## Store the ROC curve plot in another variable to display them at the end
```

```{r}
# Compute l_val for each rule.
## Store it in a variable to display at the end
```

## Results

(Grid de las ROC y las l_val)